(Re-)bridging the anatomy ontologies with SSSOM

### 2024-04-02 sssomt.tex
More corrections by N. Harris.

Building a single cross-species anatomy ontology from the ground up, with terms
suitable for all the model organisms, would be a gigantic task, requiring lots
of species-specific expertise that would be hard to gather in a single project.
Instead, a more realistic approach was adopted by the Uberon
project~\cite{mungall2012a} which, rather than attempting to describe the
anatomy of every model organism, leverages the existing species-specific
anatomy ontologies. Uberon aims to provide a ``core'' anatomy ontology that
describes anatomical structures in a species-neutral way, along with
``bridges'' that allow to integrate the more precise terms from the
species-specific ontologies (Figure~\ref{fig:bridges}). The result of merging
the Uberon core ontology with the species-specific ontologies along with the
corresponding bridges is a product called \emph{composite-metazoan} (hereafter
CM), a single, consistent cross-species ontology. The Cell Ontology (CL), an
ontology of cell types, follows the same approach~\cite{diehl2016}, and is
automatically included in Uberon's CM product.

@@

To implement its strategy, Uberon maintains a set of cross-ontology
mappings, which are used to determine where the terms from the
species-specific ontologies should be placed in the overall Uberon
hierarchy. Across ontologies of the OBO Foundry, the typical method to
represent and maintain mappings is to use cross-references (or,
informally, ``xrefs''). It’s a very simple method where, to map a term
$T_A$ in an ontology A to a term $T_B$ in a foreign ontology B, $T_A$ is
annotated with a \property{oboInOwl:hasDbXref} property, whose value is
the short identifier (``CURIE'') of $T_B$. As part of the Uberon release
pipeline, a Perl script automatically extracts those cross-references
and generates the bridge files needed to integrate the species-specific
ontologies.

While simple, the cross-reference method has several limitations. (i) It
does not allow to record any metadata about the mapping, such as: who
asserted that the two terms should be mapped? On what basis? When was
the mapping reviewed? (ii) The method does not allow for any nuance:
either two terms are mapped or they are not, without room to express
more subtle relations. (iii) Cross-references have no clear and
universally agreed upon meaning. When two terms are mapped, the meaning
of the mapping is left unspecified, and typically has to be inferred
from the ontologies that are being mapped (for example, a mapping
between the species-neutral Uberon and the \species{Drosophila}-specific
FBbt can be inferred to be a cross-species mapping). This is made worse
by the fact that cross-references in OBO ontologies are used for many
different things beyond just mappings; for example, when used to
annotate a term definition, they typically attribute a source for the
definition, akin to a citation in an academic paper.

@@

We have devised a new approach to maintain and use cross-ontology
mappings in Uberon, centered around three axes: (i) the use of a new
format to represent the mappings, (ii) the creation of dedicated mapping
relations, and (iii) the development of new tools to manipulate the
mappings and, in particular, derive OWL axioms from them.

@@

The Simple Standard for Sharing Ontological Mappings (SSSOM) is a recently
devised standard specifically intended to facilitate the exchange of semantic
mappings~\cite{matentzoglu2022b}. It allows treating mappings as first-class
data entities and to attach to them a range of metadata such as provenance and
licensing information. The standard defines a common data model to represent
mappings as well as two distinct serialisation formats to store and transport
instances of the model: a JSON-based format and a TSV-based format. The TSV
format has been specifically designed to be easily manipulatable both by
standard spreadsheet software (for editing by curators) and by common science
libraries and tools (for exploitation by data scientists and engineers). At
least two independent implementations of the standard are available: SSSOM-Py
and SSSOM-Java, for the Python and Java programming languages, respectively. In
addition, several ontology-related tools have started to add direct support for
the format, such as the Ontology Access Kit~\cite{mungall2023} and the Ontology
Development Kit~\cite{matentzoglu2022a}. The community resource Biomappings
also provides its mappings in SSSOM, among other formats~\cite{hoyt2023}.

@@

One of the benefits brought by the SSSOM standard is the possibility to use
highly specific mapping predicates to express precisely what is the intended
meaning of a mapping. While many mapping sets don’t actually need this, and can
use common mapping predicates such as those from the SKOS
vocabulary~\cite{bechhofer2009} (\property{skos:exactMatch},
\property{skos:narrowMatch}, etc.), cross-species mappings are an example of an
application where the common predicates are not satisfying. For example, let us
consider a mapping between the FBbt term for neuron (\curie{FBbt:00005106},
representing a neuron in a fruit fly) and the corresponding term in the Cell
Ontology (\curie{CL:0000540}, representing a neuron in any species): clearly it
would not be correct to state that a fly neuron is the same concept as a neuron
in any other species --~the two terms are not interchangeable, so
\property{skos:exactMatch} is not a suitable mapping predicate.  A slightly
more correct predicate could be \property{skos:narrowMatch}, because is it true
that a fly neuron is a narrower concept than a species-neutral neuron; but a
sensory neuron (\curie{CL:0000101}) is also a narrower concept than a generic
neuron, yet the relation between \term{sensory neuron} and \term{neuron} is of
a different nature than the relation between \term{fly neuron} and
\term{(species-neutral) neuron}.

Overall we believe that cross-species mappings are \emph{sui generis}
mappings and that they warrant the use of dedicated mapping predicates
to reflect that fact. We therefore expanded the SEMAPV vocabulary,
specifically established for use with SSSOM~\cite{matentzoglu2022}, with
four new mapping predicates (Figure~\ref{fig:relations}) that mirrors
existing SKOS predicates but are specifically intended for cases where
the subject and object of a mapping belong to different taxonomic
groupings.

@@

Even with fast SSSOM adoption, it is likely that many tools will remain
unable to directly exploit SSSOM sets in the foreseeable future.
Therefore, when mappings are stored in the SSSOM format, we need to be
able to convert them into OWL axioms that can be used by any ontology
manipulation tool. To that effect, we have developed a plugin for ROBOT,
the standard tool used to manipulate OBO ontologies~\cite{jackson2019}.
The plugin is built on top of the Java implementation of the SSSOM
standard and provides a \emph{sssom:inject} command that takes a SSSOM
mapping set as input, derives OWL axioms from the mappings, and injects
them into an ontology.

Rather than hardcoding the logic for deriving the axioms --~which would
have been efficient but would have made the logic harder to modify and
re-use~--, we designed a small domain-specific language (DSL), named
SSSOM/T-OWL (``SSSOM/Transform-to-OWL''), that allows users to describe
which mappings in a set should be transformed into OWL axioms and what
kind of axioms should be produced.

@@

Filters allow to compare the value of any of the SSSOM metadata slots
with a target value and to discard any mapping with a different value. A
basic form of wildcard matching is supported, where a filter can select
mappings with a value that starts with the same prefix as the target
value. Individual filters can be combined to select mappings based on
more than just one metadata slot; they can also be negated, so that a
filter that would select a given set of mappings would, after negation,
discard that set and instead select the complementary set of mappings.

Preprocessors allow modifying a mapping before it is used by subsequent
rules. Possible modifications include inverting a mapping (the subject
becomes the object and vice-versa) and changing some of the mapping's
metadata. A preprocessor can also be used to completely remove a mapping
from the set, so that no further rule will be applied to this mapping.

Finally, the generators take the selected mapping and turn it into an axiom.
The most important generator function is \function{create\_axiom}, which can
produce an arbitrary axiom described by an expression in Manchester
syntax~\cite{patel-schneider2012}. Other generators allow to annotate classes
of the ontology with any of the available mappings metadata.

Of note, the current implementation of the SSSOM/T-OWL language clearly
separates the OWL generators from the filters and the preprocessing
functions. This makes it easy to create other domain-specific languages
similar to SSSOM/T-OWL, using the same overall syntax, the same filters
and the same preprocessors, but that generate other objects than OWL
axioms.

@@

Mappings between Uberon and the other species-specific anatomy
ontologies beyond FBbt are still currently maintained as
cross-references either within Uberon or within the species-specific
ontologies. They are extracted and converted to SSSOM dynamically as
part of Uberon's release pipeline. In a second phase, which we have not
started yet, we will migrate those mappings to SSSOM definitely, as was
done for the FBbt mappings.

@@

Regardless of their origin, all mappings between Uberon, CL, and the
species-specific ontologies are then processed by a single large
SSSOM/T-OWL ruleset that contains all the required logic to generate the
bridging axioms. A typical rule in this ruleset is the following:

@@

All axioms produced by the SSSOM/T-OWL ruleset are dispatched into one
different bridge file for each foreign ontology.

@@

Once the bridge files have been generated, an intermediate ontology
(\emph{collected-metazoan}) is built by merging together: Uberon itself,
the Cell Ontology, all the species-specific ontologies, and their
corresponding bridges. Finally, \emph{composite-metazoan} itself is
derived from this intermediate product by applying some custom logic
(implemented in a small, dedicated ROBOT plugin) to ``prune'' redundant
terms and replace them by equivalent anonymous class expressions.

@@

The ``pruning'' operation leading to \emph{composite-metazoan} consists
in replacing all occurrences of \curie{FBbt:00004865} by the anonymous
expression it is equivalent to. So the following axiom, which states
that the \term{oviduct} (\curie{FBbt:00004911}) is \term{continuous
with} (\curie{RO:0002150}) the fly ovary:

@@

While full transition to SSSOM is not achieved yet, we now have the
needed infrastructure in place to complete this migration, and we have a
more robust and more flexible pipeline for creating the bridges needed
for \emph{composite-metazoan}. The logic for deriving the bridges is
expressed using a simple DSL that most editors should be able to master
quickly, should the need arise to modify said logic (for example to add
a new bridge). While the precise SSSOM/T-OWL rules we use are obviously
specific to the needs of Uberon, we believe the general approach of
maintaining mappings as SSSOM sets and transforming them into OWL
bridges using a simple DSL could be re-used by any project that requires
integrating several ontologies together.

>>>>

Building a single cross-species anatomy ontology from the ground up,
with terms suitable for all the model organisms, would be a gigantic
task, requiring lots of species-specific expertise that would be hard to
gather in a single project.  Instead, a more practical approach was
adopted by the Uberon project~\cite{mungall2012a} which, rather than
attempting to describe the anatomy of every model organism, leverages
the existing species-specific anatomy ontologies. Uberon aims to provide
a ``core'' anatomy ontology that describes anatomical structures in a
species-neutral way, along with ``bridges'' that allow to integrate the
more precise terms from the species-specific ontologies
(Figure~\ref{fig:bridges}). The result of merging the Uberon core
ontology with the species-specific ontologies along with the
corresponding bridges is a product called \emph{composite-metazoan}
(hereafter CM), a single, consistent cross-species ontology. The Cell
Ontology (CL), an ontology of cell types, follows the same
approach~\cite{diehl2016}, and is automatically included in Uberon's CM
product.

@@

To implement this strategy, Uberon maintains a set of cross-ontology
mappings, which are used to determine where the terms from the
species-specific ontologies should be placed in the overall Uberon
hierarchy. Across ontologies of the OBO Foundry, the typical method to
represent and maintain mappings is to use cross-references (or,
informally, ``xrefs''). It’s a very simple method where, to map a term
$T_A$ in an ontology A to a term $T_B$ in a foreign ontology B, $T_A$ is
annotated with a \property{oboInOwl:hasDbXref} property, whose value is
the short identifier (``CURIE'') of $T_B$. As part of the Uberon release
pipeline, a Perl script automatically extracts those cross-references
and generates the bridge files needed to integrate the species-specific
ontologies.

While simple, the cross-reference method has several limitations. (i) It
does not allow to record any metadata about the mapping, such as: who
asserted that the two terms should be mapped? On what basis? When was
the mapping reviewed?  (ii) The method does not allow for any nuance:
either two terms are mapped or they are not, without room to express
more subtle relations. (iii) Cross-references have no clear and
universally agreed upon meaning. When two terms are mapped to each
other, the meaning of the mapping is left unspecified, and typically has
to be inferred from the ontologies that are being mapped (for example, a
mapping between the species-neutral Uberon and the
\species{Drosophila}-specific FBbt can be inferred to be a cross-species
mapping). This is made worse by the fact that cross-references in OBO
ontologies are used for many different things beyond just mappings; for
example, when used to annotate a term definition, cross-references
typically attribute a source for the definition, akin to a citation in
an academic paper.

@@

We have devised a new approach to maintain and use cross-ontology
mappings in Uberon, centred around three axes: (i) the use of a new
format to represent the mappings, (ii) the creation of dedicated mapping
relations, and (iii) the development of new tools to manipulate the
mappings and, in particular, derive OWL axioms from them.

@@

The Simple Standard for Sharing Ontological Mappings (SSSOM) is a
recently devised standard specifically intended to facilitate the
exchange of semantic mappings~\cite{matentzoglu2022b}. It allows
mappings to be treated as first-class data entities and to attach to
them a range of metadata such as provenance and licensing information.
The standard defines a common data model to represent mappings as well
as two distinct serialisation formats to store and transport instances
of the model: a JSON-based format and a TSV-based format. The TSV format
has been specifically designed to be easily manipulatable both by
standard spreadsheet software (for editing by curators) and by common
science libraries and tools (for use by data scientists and engineers).
At least two independent implementations of the standard are available:
SSSOM-Py and SSSOM-Java, for the Python and Java programming languages,
respectively. In addition, several ontology-related tools have started
to add direct support for the format, such as the Ontology Access
Kit~\cite{mungall2023} and the Ontology Development
Kit~\cite{matentzoglu2022a}. The community resource Biomappings also
provides its mappings in SSSOM, among other formats~\cite{hoyt2023}.

@@

One of the benefits brought by the SSSOM standard is the possibility to
use highly specific mapping predicates to express precisely the intended
meaning of a mapping. While many mapping sets don’t actually need this,
and can use common mapping predicates such as those from the SKOS
vocabulary~\cite{bechhofer2009} (\property{skos:exactMatch},
\property{skos:narrowMatch}, etc.), cross-species mappings are an
example of an application where the common predicates are not
sufficient. For example, let us consider a mapping between the FBbt term
for neuron (\curie{FBbt:00005106}, representing a neuron in a fruit fly)
and the corresponding term in the Cell Ontology (\curie{CL:0000540},
representing a neuron in any species): clearly it would not be correct
to state that a fly neuron is the same concept as a neuron in any other
species --~the two terms are not interchangeable, so
\property{skos:exactMatch} is not a suitable mapping predicate.  A
slightly more correct predicate could be \property{skos:narrowMatch},
because is it true that a fly neuron is a narrower concept than a
species-neutral neuron; but a sensory neuron (\curie{CL:0000101}) is
also a narrower concept than a generic neuron, yet the relation between
\term{sensory neuron} and \term{neuron} is of a different nature than
the relation between \term{fly neuron} and \term{(species-neutral)
neuron}.

Overall we believe that cross-species mappings are \emph{sui generis}
mappings and that they warrant the use of dedicated mapping predicates
to reflect that fact. We therefore expanded the SEMAPV vocabulary,
specifically established for use with SSSOM~\cite{matentzoglu2022}, with
four new mapping predicates (Figure~\ref{fig:relations}) that mirror
existing SKOS predicates but are specifically intended for cases where
the subject and object of a mapping belong to different taxonomic
groupings.

@@

Even as SSSOM adoption spreads rapidly, it is likely that many tools
will remain unable to directly exploit SSSOM sets in the foreseeable
future.  Therefore, when mappings are stored in the SSSOM format, we
need to be able to convert them into OWL axioms that can be used by any
ontology manipulation tool. To that effect, we have developed a SSSOM
plugin for ROBOT, the standard tool used to manipulate OBO
ontologies~\cite{jackson2019}.  The plugin is built on top of the Java
implementation of the SSSOM standard and provides a \emph{sssom:inject}
command that takes a SSSOM mapping set as input, derives OWL axioms from
the mappings, and injects them into an ontology.

Rather than hardcoding the logic for deriving the axioms --~which would
have been efficient but would have made the logic harder to modify and
re-use~-- we designed a small domain-specific language (DSL), named
SSSOM/T-OWL (``SSSOM/Transform-to-OWL''), that allows users to describe
which mappings in a set should be transformed into OWL axioms and what
kind of axioms should be produced.

@@

Filters allow the value of any of the SSSOM metadata slots to be
compared with a target value and to discard any mapping with a different
value. A basic form of wildcard matching is supported, where a filter
can select mappings with a value that starts with the same prefix as the
target value. Individual filters can be combined to select mappings
based on more than just one metadata slot; they can also be negated, so
that a filter that would select a given set of mappings would, after
negation, instead select the complementary set of mappings.

Preprocessors allow a mapping to be modified before it is used by
subsequent rules. Possible modifications include inverting a mapping
(the subject becomes the object and vice-versa) and changing some of the
mapping's metadata. A preprocessor can also be used to completely remove
a mapping from the set, so that no further rule will be applied to this
mapping.

Finally, the generators take the selected mapping and turn it into an
axiom.  The most important generator function is
\function{create\_axiom}, which can produce an arbitrary axiom described
by an expression in Manchester syntax~\cite{patel-schneider2012}. Other
generators allow classes of the ontology to be annotated with any of the
available mappings metadata.

Of note, the current implementation of the SSSOM/T-OWL language cleanly
separates the OWL generators from the filters and the preprocessing
functions.  This makes it easy to create other domain-specific languages
similar to SSSOM/T-OWL, using the same overall syntax, the same filters
and the same preprocessors, but that generate objects other than OWL
axioms.

@@

Mappings between Uberon and the other species-specific anatomy
ontologies beyond FBbt are still currently maintained as
cross-references either within Uberon or within the species-specific
ontologies. They are extracted and converted to SSSOM dynamically as
part of Uberon's release pipeline. In a second phase, which we have not
started yet, we will migrate those mappings fully to SSSOM, as was done
for the FBbt mappings.

@@

Regardless of their origin, all mappings between Uberon, CL, and the
species-specific ontologies are processed by a single large SSSOM/T-OWL
ruleset that contains all the required logic to generate the bridging
axioms. A typical rule in this ruleset is the following:

@@

All axioms produced by the SSSOM/T-OWL ruleset are saved into a separate
bridge file for each foreign ontology.

@@

Once the bridge files have been generated, an intermediate ontology
(\emph{collected-metazoan}) is built by merging together Uberon itself,
the Cell Ontology, all the species-specific ontologies, and their
corresponding bridges. Finally, \emph{composite-metazoan} itself is
derived from this intermediate product by applying some custom logic
(implemented in a small, dedicated ROBOT plugin) to ``prune'' redundant
terms and replace them with equivalent anonymous class expressions.

@@

The ``pruning'' operation leading to \emph{composite-metazoan} consists
of replacing all occurrences of \curie{FBbt:00004865} by the anonymous
expression it is equivalent to. So the following axiom, which states
that the \term{oviduct} (\curie{FBbt:00004911}) is \term{continuous
with} (\curie{RO:0002150}) the fly ovary:

@@

To construct a consistent and maintainable cross-species anatomy
ontology, we have updated the integration mechanism between Uberon, CL,
and the species-centric ontologies by using SSSOM to manage the mappings
betweeen the ontologies. While full transition to SSSOM is not yet
complete, we now have the needed infrastructure in place to complete
this migration, and we have a more robust and more flexible pipeline for
creating the bridges needed for \emph{composite-metazoan}. The logic for
deriving the bridges is expressed using a simple DSL (SSSOM/T-OWL) that
most ontology editors should be able to master quickly, should the need
arise to modify said logic (for example to add a new bridge). While the
precise SSSOM/T-OWL rules we use are obviously specific to the needs of
Uberon, we believe the general approach of maintaining mappings as SSSOM
sets and transforming them into OWL bridges using a simple DSL could be
used by any project that requires integrating several ontologies
together.
